PyUltraLight: A Pseudo-Spectral Solver for Ultralight Dark Matter Dynamics

### 2018-09-29 draft.tex
Responses to RE's comments

\usepackage[T1]{fontenc} % if needed
\usepackage{eucal}
\usepackage{xspace}
\usepackage{color}
\usepackage{url}
\newcommand{\re}[1]{\textcolor{blue}{[{\bf RE}: #1]}}

@@

\PyUltraLight is based on a symmetrised-split-step (leapfrog) solver for the time evolution, and uses a pseudospectral Fourier algorithm to solve the Poisson equation for the gravitational potential at each step.\footnote{A similar methodology was described in Ref.~\cite{Paredes:2015wga}; to our knowledge this code has not been released. Spectral methods are often used to solve the Poisson equation in large scale structure simulations, while the {\sc PSpectre} code \cite{Easther:2010qz} provides a pseudospectral solver for the evolution of the inflaton  and fields coupled to it during parametric resonance and preheating after inflation \cite{Amin:2010dc,Amin:2011hj,Zhou:2013tsa}.} This algorithm has $2^{nd}$ order accurate time integration steps and sub-percent level energy conservation, while the wavefunction normalisation  is conserved to machine precision. As a pseudospectral code, linear differential operators are computed by direct multiplication in the Fourier domain, while non-linear terms are evaluated in position space. Consequently, \PyUltraLight is free from noise associated with spatial derivatives computed via finite-differencing. There is a necessary computational cost associated with the Fourier and inverse Fourier transforms but these transforms are optimised in \PyUltraLight through the use of the pyFFTW pythonic wrapper around the C-based FFTW subroutine library \cite{pyfftw,fftw}. As the FFTW libraries offer full parallelisation, \PyUltraLight is currently designed to take advantage of multiple cores on a user PC or shared-memory environment. Full MPI compatibility has not yet been implemented as we have not found a need to run simulations in a distributed-memory cluster environment, however future releases may address this possibility.

@@

    \nabla^2\Phi=4\pi G m \big(\vert\psi\vert^2-\langle\vert\psi\vert^2\rangle\big).

@@

\re{do we define $\langle \rangle$? Also -- should we explain in a sentence WHY the form of the equation changes when we go to period b.c.s?}

@@

The algorithm can be represented schematically as
\begin{equation}\label{eq:schematic-psi}
    \psi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)=\operatorname{exp}\left[-\frac{ih}{2}\Phi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)\right]\CMcal{F}^{-1} \operatorname{exp}\left[\frac{-ih}{2}k^2\right]\CMcal{F}\operatorname{exp}\left[-\frac{ih}{2}\Phi(\vec{x},t)\right]\psi(\vec{x},t),
\end{equation}
where the order of operations runs from right to left, $\CMcal{F}$ and $\CMcal{F}^{-1}$ denote the discrete Fourier transform and its inverse, and $k$ is the wavenumber in the Fourier domain. The potential field is updated following the inverse Fourier transform in equation \ref{eq:schematic-psi}, via
%
\begin{equation}\label{eq:schematic-phi}
    \Phi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)=\CMcal{F}^{-1}\left(-\frac{1}{k^2}\right)\CMcal{F}\ 4\pi\vert\psi(\vec{x},t_{i})\vert^2,
\end{equation}
%
where $\psi(\vec{x},t_{i})$ is the field configuration at this halfway point in the full timestep. \re{We explicitly set the $k=0$ Fourier mode to zero prior to the final inverse Fourier transform; as a consequence there is no need to subtract the global average density from the local value in Equation \ref{eq:schematic-phi}, in contrast to Equation \ref{eq:poisson_periodic}.} The final operation in equation \ref{eq:schematic-psi}  only changes the phase of  $\psi$, so we could replace $\psi(\vec{x},t_{i})$ with $\psi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)$ in equation \ref{eq:schematic-phi} with no change in meaning. \PyUltraLight makes an additional simplification to the symmetrised split-step Fourier method by combining the consecutive half-steps in the nonlinear term into a single full step. Consequently, only the first and last operations involve actual half steps. Schematically this becomes 
%
\begin{equation}
    \psi(t\hspace*{-.2em}+\hspace*{-.2em}nh)=\operatorname{exp}\left[+\frac{ih}{2}\Phi\right]\left(\prod^n\operatorname{exp}\left[-ih\Phi\right]\operatorname{exp}\left[\frac{ih}{2}\nabla^2\right]\right)\operatorname{exp}\left[-\frac{ih}{2}\Phi\right]\psi(t),
\end{equation}
%
where  $\Phi$  is updated at each step via equation \ref{eq:schematic-phi}; attention is drawn to the sign difference between the first and last operators.

@@

The default timestep in  \PyUltraLight  is fixed with reference to the fluid interpretation of the Schr{\"o}dinger-Poisson system \re{CITATION}. We obtain this via
%
\begin{equation}\label{eq:madelung}
    \psi\equiv\sqrt{\rho}e^{i\theta}, \quad \vec{v}\equiv\boldsymbol{\nabla}\theta
\end{equation}
%
\re{Do we need a bit more detail, or a reference?}  If the phase change between two adjacent grid points across a timestep, exceeds $\pi$ the fluid will appear to move backwards across the grid, and we set the default timestep, $\Delta t$, such that fluid travelling at this maximum velocity will traverse one grid space, $\Delta x$, per timestep, or
\begin{equation}\label{eq:timestep}
    \Delta t = \frac{(\Delta x)^2}{\pi}.
\end{equation}
%
This is  a choice, rather than a strict constraint but if the ``fluid'' approaches velocities where the phase appears to switch direction, the configuration will also be approaching the point where the simulation grid is too coarse to fully resolve the dynamics. Making the timestep much smaller than this value may offer little practical advantage. Conversely, in some cases the breakdown may occur in regions of the simulation volume of little physical interest, and the user is free to choose a larger timestep  via the 'step\textunderscore factor' parameter in the Jupyter notebook.

Alternatively, Ref.~\cite{Mocz:2017wlg} fixes the timestep by ensuring that neither of the unitary operators in Equation \ref{eq:schematic-psi}  leads to a phase change of more than $\pi$ for a single grid point over one timestep. However, because the pseudo-spectral algorithm does not compare the phase of a single gridpoint at different points in time, this choice of timestep is  not a requirement for stability. This method  give the following constraints: 
\begin{equation}
    \Delta t < \left[\frac{2\pi}{\Phi_{max}}, \frac{2(\Delta x)^2}{\pi}\right],
\end{equation}
%
where the second of these constraints is generally the stricter of the two, and is equivalent to the default choice up to a multiplicative factor of $\mathcal{O}(1)$. Our experience is that specifying the timestep via Equation \ref{eq:timestep} is suitable for the majority of simulation scenarios, and we explore convergence in more detail in Section \ref{sec:resolution}.

@@

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 0 0 0,clip]{first_and_last}
  \caption{The configuration used to test the sensitivity of solutions to the spatial and temporal resolution. Two solitons of mass $m=20$ are initialised at radial distances $r=2$ from a central mass with $M=1000$ moving in opposite directions with initial speeds $\vert v\vert=\sqrt{M/r}$, corresponding to clockwise orbits around the central mass. The  box size is 10, while the total duration is 0.25 (all quantities in code units. Time runs from left to right. \re{you could actually zoom in, to only include the active region.} }
  \label{fig:disruption}
\end{figure}

To examine the sensitivity of the  $\psi$ field configuration to the spatial resolution, we first run at $256^3$ with the default timestep. We then re-run  at resolutions  from $64^3$ to $320^3$ with the timestep fixed to the  $256^3$ value and downsample the final outputs to $64^3$. We  sort the resulting values by  the density at the corresponding spatial location, and plot differences in the phase and the magnitude of $\psi$ relative to the values of the $320^3$ run in Figures \ref{fig:spatial_magnitudes} and \ref{fig:spatial_phases}. \re{maybe this could be one combined figure with both plots.}  The convergence is poor  at $\psi$ at $64^3$, but improves with resolution, to the point that there is little difference between the  $256^3$ and $320^3$ cases.

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 1cm 0 1.2cm,clip]{spatial_discretisation}
  \caption{Improving convergence of $\vert\psi\vert$ with increased spatial resolution for the simulation shown in Figure \ref{fig:disruption}. Field values are arranged in order of increasing magnitude of the highest resolution run from left to right.}
  \label{fig:spatial_magnitudes}
\end{figure}

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 0.5cm 0 1.2cm,clip]{phases}
  \caption{Deviation of the phase of $\psi$ compared to the highest resolution result ($320^3$). Field values are arranged in order of increasing magnitude from left to right. A slight improvement in phase convergence can be seen for higher density regions to the right.}
  \label{fig:spatial_phases}
\end{figure}

To examine the sensitivity of the the  $\psi$ field configuration to the timestep, we take the same default simulation at $256^3$, and then compare this to runs with timesteps 0.1, 10, and 50 times the default and down-sample the final output arrays to $64^3$. We sort sort the array values in order of the $\psi$ field magnitude in the run with the smallest timestep and in Figures \ref{fig:temporal_magnitudes} and \ref{fig:temporal_phases} \re{maybe also combined into one full page figure?} we show the difference in the phase and magnitude of $\psi$ as a function of the timestep.  The difference between the results with the default timestep and a value 10 times smaller are negligible; and there is reasonable agreement between the default case and those with the timestep boosted by a factor of 10. However, when the timestep is increased by a factor of $50$ the accuracy of both the phase and magnitude data are significantly reduced.

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 1cm 0 1cm,clip]{temporal_discretisation}
  \caption{Deviation of the magnitude of $\psi$ as compared to the simulation run with 0.1 times the default timestep. The convergence of the results for the default timestep with the results from the smaller timestep run demonstrate the suitability of the chosen default. We see relatively good convergence, particularly in high density regions, for a $<10$-fold increase above the default timestep.}
  \label{fig:temporal_magnitudes}
\end{figure}

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 0.5cm 0 1cm,clip]{temporal_phases}
  \caption{Deviation of the phase of $\psi$ from the results of the simulation in which the timestep is set to 0.1 times the default. Again we see excellent convergence for the default timestep, and reasonable convergence for $<10$ times the default, with accuracy improving at high density regions.}
  \label{fig:temporal_phases}
\end{figure}

>>>>

\usepackage[T1]{fontenc} % if needed
\usepackage{eucal}
\usepackage{xspace}
\usepackage{color}
\usepackage{url}
\newcommand{\re}[1]{\textcolor{blue}{[{\bf RE}: #1]}}
\newcommand{\ek}[1]{\textcolor{red}{[{\bf EK}: #1]}}

@@

\PyUltraLight is based on a symmetrised-split-step (leapfrog) solver for the time evolution, and uses a pseudospectral Fourier algorithm to solve the Poisson equation for the gravitational potential at each step.\footnote{A similar methodology was described in Ref.~\cite{Paredes:2015wga}; at the time of writing this code has not been released. Spectral methods are often used to solve the Poisson equation in large scale structure simulations, while the {\sc PSpectre} code \cite{Easther:2010qz} provides a pseudospectral solver for the evolution of the inflaton  and fields coupled to it during parametric resonance and preheating after inflation \cite{Amin:2010dc,Amin:2011hj,Zhou:2013tsa}.} This algorithm has $2^{nd}$ order accurate time integration steps and sub-percent level energy conservation, while the wavefunction normalisation  is conserved to machine precision. As a pseudospectral code, linear differential operators are computed by direct multiplication in the Fourier domain, while non-linear terms are evaluated in position space. Consequently, \PyUltraLight is free from noise associated with spatial derivatives computed via finite-differencing. There is a necessary computational cost associated with the Fourier and inverse Fourier transforms but these transforms are optimised in \PyUltraLight through the use of the pyFFTW pythonic wrapper around the C-based FFTW subroutine library \cite{pyfftw,fftw}. As the FFTW libraries offer full parallelisation, \PyUltraLight is currently designed to take advantage of multiple cores on a user PC or shared-memory environment. Full MPI compatibility has not yet been implemented as we have not found a need to run simulations in a distributed-memory cluster environment, however future releases may address this possibility.

@@

    \nabla^2\Phi=4\pi G m \big(\vert\psi\vert^2-\langle\vert\psi\vert^2\rangle\big),

@@

where we subtract the average density from the right hand side of the Poisson equation. The form of Equation \ref{eq:poisson_periodic} is a consequence of Gauss' law and the fact that the surface integral of the gradient of the field is identically zero when periodic boundary conditions are imposed.

\re{do we define $\langle \rangle$? Also -- should we explain in a sentence WHY the form of the equation changes when we go to period b.c.s?}\ek{I've given slightly more info here, think this is sufficient to understand the form of 2.8}

@@

The algorithm can be represented schematically as
\begin{equation}\label{eq:schematic-psi}
    \psi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)=\operatorname{exp}\left[-\frac{ih}{2}\Phi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)\right]\CMcal{F}^{-1} \operatorname{exp}\left[\frac{-ih}{2}k^2\right]\CMcal{F}\operatorname{exp}\left[-\frac{ih}{2}\Phi(\vec{x},t)\right]\psi(\vec{x},t),
\end{equation}
where the order of operations runs from right to left, $\CMcal{F}$ and $\CMcal{F}^{-1}$ denote the discrete Fourier transform and its inverse, and $k$ is the wavenumber in the Fourier domain. The potential field is updated following the inverse Fourier transform in equation \ref{eq:schematic-psi}, via
%
\begin{equation}\label{eq:schematic-phi}
    \Phi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)=\CMcal{F}^{-1}\left(-\frac{1}{k^2}\right)\CMcal{F}\ 4\pi\vert\psi(\vec{x},t_{i})\vert^2,
\end{equation}
%
where $\psi(\vec{x},t_{i})$ is the field configuration at this halfway point in the full timestep. We explicitly set the $k=0$ Fourier mode to zero prior to the final inverse Fourier transform; as a consequence there is no need to subtract the global average density from the local value in Equation \ref{eq:schematic-phi}, in contrast to Equation \ref{eq:poisson_periodic}.\ek{Incorporated this explanation into main text} The final operation in equation \ref{eq:schematic-psi}  only changes the phase of  $\psi$, so we could replace $\psi(\vec{x},t_{i})$ with $\psi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)$ in equation \ref{eq:schematic-phi} with no change in meaning. \PyUltraLight makes an additional simplification to the symmetrised split-step Fourier method by combining the consecutive half-steps in the nonlinear term into a single full step. Consequently, only the first and last operations involve actual half steps. Schematically this becomes 
%
\begin{equation}
    \psi(t\hspace*{-.2em}+\hspace*{-.2em}nh)=\operatorname{exp}\left[+\frac{ih}{2}\Phi\right]\left(\prod^n\operatorname{exp}\left[-ih\Phi\right]\operatorname{exp}\left[\frac{ih}{2}\nabla^2\right]\right)\operatorname{exp}\left[-\frac{ih}{2}\Phi\right]\psi(t),
\end{equation}
%
where  $\Phi$  is updated at each step via equation \ref{eq:schematic-phi}; attention is drawn to the sign difference between the first and last operators.

@@

The default timestep in  \PyUltraLight  is fixed with reference to the fluid interpretation of the Schr{\"o}dinger-Poisson system \cite{Hui:2016ltb}. The fluid interpretation is often used to recast the Schr{\"o}dinger-Poisson system in the form of the Madelung equations \cite{Suarez:2011yf}, such that the system can then be studied using the principles of hydrodynamics.\ek{Slightly more info and reference given here, though I don't want to go into too much detail as the specifics of the fluid representation aren't needed here.} The first step is to define:
\begin{equation}\label{eq:madelung}
    \psi\equiv\sqrt{\rho}e^{i\theta}, \quad \vec{v}\equiv\boldsymbol{\nabla}\theta.
\end{equation}
From this definition of the fluid velocity, we see that if the phase change between two adjacent grid points exceeds $\pi$,\ek{not over a timestep, but at the same point in time} the fluid will appear to move backwards across the grid. Therefore, we set the default timestep, $\Delta t$, such that fluid travelling at this maximum velocity will traverse one grid space, $\Delta x$, per timestep, or
\begin{equation}\label{eq:timestep}
    \Delta t = \frac{(\Delta x)^2}{\pi}.
\end{equation}
This is a choice, rather than a strict constraint on the timestep. We motivate this choice of timestep by recognising that if the ``fluid'' approaches velocities where the phase appears to switch direction, the configuration will also be approaching the point where the simulation grid is too coarse to fully resolve the dynamics. Hence, making the timestep much smaller than this value may offer little practical advantage. Conversely, in some cases the breakdown may occur in regions of the simulation volume of little physical interest, and the user is free to choose a larger timestep  via the 'step\textunderscore factor' parameter in the Jupyter notebook.

Alternatively, Ref.~\cite{Mocz:2017wlg} fixes the timestep by ensuring that neither of the unitary operators in Equation \ref{eq:schematic-psi}  lead to a phase change of more than $\pi$ for a single grid point over one timestep. However, because the pseudo-spectral algorithm does not compare the phase of a single gridpoint at different points in time, this choice of timestep is not a requirement for stability. This method gives the following constraints: 
\begin{equation}
    \Delta t < \left[\frac{2\pi}{\Phi_{max}}, \frac{2(\Delta x)^2}{\pi}\right],
\end{equation}
%
where the second of these constraints is generally the stricter of the two, and is equivalent to our default choice of timestep up to a multiplicative factor of $\mathcal{O}(1)$. Our experience is that specifying the timestep via Equation \ref{eq:timestep} is suitable for the majority of simulation scenarios, and we explore convergence in more detail in Section \ref{sec:resolution}.

@@

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 0 0 0,clip]{first_and_last}
  \caption{The configuration used to test the sensitivity of solutions to the spatial and temporal resolution. Two solitons of mass $m=20$ are initialised at radial distances $r=2$ from a central mass with $M=1000$ moving in opposite directions with initial speeds $\vert v\vert=\sqrt{M/r}$, corresponding to clockwise orbits around the central mass. The  box size is 10, while the total duration is 0.25 (all quantities in code units. Time runs from left to right. \re{you could actually zoom in, to only include the active region.}\ek{I'm just concerned about misrepesenting the size of the simulation grid - I wanted to accurately represent the buffer zone, because insufficient buffer will affect the dynamics, and people trying to reproduce these results might get confused. In any case, the visual details of the density distribution aren't important here, the purpose of the figure is just to give an idea of the setup} }
  \label{fig:disruption}
\end{figure}

To examine the sensitivity of the  $\psi$ field configuration to the spatial resolution, we first run at $256^3$ with the default timestep. We then re-run  at resolutions  from $64^3$ to $320^3$ with the timestep fixed to the  $256^3$ value and downsample the final outputs to $64^3$. We  sort the resulting values by  the density at the corresponding spatial location, and plot differences in the phase and the magnitude of $\psi$ relative to the values of the $320^3$ run as shown in Figure \ref{fig:spatial} (bottom). \re{maybe this could be one combined figure with both plots.}\ek{Done.} The convergence is poor at $64^3$, but improves with resolution, to the point that there is little difference between the  $256^3$ and $320^3$ cases.

To examine the sensitivity of the the  $\psi$ field configuration to the timestep, we take the same default simulation at $256^3$, and then compare this to runs with timesteps 0.1, 10, and 50 times the default and down-sample the final output arrays to $64^3$. We sort sort the array values in order of the $\psi$ field magnitude in the run with the smallest timestep and in Figure \ref{fig:temporal} \re{maybe also combined into one full page figure?}\ek{Done.} we show the difference in the phase and magnitude of $\psi$ as a function of the timestep.  The difference between the results with the default timestep and a value 10 times smaller are negligible; and there is reasonable agreement between the default case and those with the timestep boosted by a factor of 10. However, when the timestep is increased by a factor of $50$ the accuracy of both the phase and magnitude data are significantly reduced.

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 1cm 0 2cm,clip]{combined2}
  \caption{Top: Deviation of the phase of $\psi$ compared to the highest resolution result ($320^3$). Field values are arranged in order of increasing magnitude from left to right. A slight improvement in phase convergence can be seen for higher density regions to the right. Bottom: Improving convergence of $\vert\psi\vert$ with increased spatial resolution for the simulation shown in Figure \ref{fig:disruption}}
  \label{fig:spatial}
\end{figure}

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 1cm 0 2cm,clip]{combined}
  \caption{Top: Deviation of the phase of $\psi$ from the results of the simulation in which the timestep is set to 0.1 times the default. We see excellent convergence for the default timestep, and reasonable convergence for $<10$ times the default, with accuracy improving at high density regions. Bottom: Deviation of the magnitude of $\psi$ as compared to the simulation run with 0.1 times the default timestep. The convergence of the results for the default timestep with the results from the smaller timestep run demonstrate the suitability of the chosen default. We see relatively good convergence, particularly in high density regions, for a $<10$-fold increase above the default timestep.}
  \label{fig:temporal}
\end{figure}

\ek{added new content from here to the end of the section}
To supplement our analysis of the dependence of the field configuration on spatial and temporal resolutions, we include Figures \ref{fig:comparison} and \ref{fig:comparison2}. These figures represent the density profiles down the axis of symmetry of the initial configuration (vertical axis in Figure \ref{fig:disruption}), however rather than the profiles at $t=0$, these are the profiles after approximately half a revolution around the central potential at t=0.28 code units. At this point, the two initial solitons have become distorted due to tidal forces, but have yet to become completely disrupted. Figure \ref{fig:comparison} demonstrates the effect of increasing the timestep from 0.1 times the default value to a maximum of 50 times the default value. We see that the profiles for the default timestep and 0.1 times the default timestep are virtually indistinguishable, with 10 times the default timestep performing resonably well, with small deviations at high densities. Meanwhile, at 50 times the default timestep the results are significantly distorted. Figure \ref{fig:comparison2} demonstrates the effect of decreasing the spatial resolution from $320^3$ to $64^3$. As expected, the lowest resolution performs poorly, whereas we see high convergence of the density profile at resolutions above $192^3$.

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 0.8cm 0 1.2cm,clip]{comparison}
  \caption{Effect of increasing the timestep on the density profile at half a revolution.}
  \label{fig:comparison}
\end{figure}

\begin{figure}
  \includegraphics[width=1.\textwidth,trim=0 0.8cm 0 1.2cm,clip]{comparison2}
  \caption{Effect of decreasing the spatial resolution on the density profile at half a revolution.}
  \label{fig:comparison2}
\end{figure}
